Best practices for forecasting changes in biodiversity: an example using breeding birds

### 2017-08-26 manuscript/draft.Rmd

![Performance of the six models, trained with data from 1982-2003, when 
predicting one year into the future (2004) or ten years into the future (2013).
In general, the single-site models (**A.**) outperformed the environmental
models (**B.**). The accuracy of the predictions generally declined as the
timescale of the forecast was extended from 2004 to 2013.](../figures/scatter.png)

While the models generally did well in absolute terms (Figure 3), none
consistently outperformed the "average" baseline (Figure 4). The auto-ARIMA was
generally the best-performing non-baseline model, but in many cases (67% of the
time), the auto.arima procedure automatically selected a model with only an
intercept term (i.e. no autoregressive terms, no drift, and no moving average
terms), making it similar to the "average" model. All five alternatives to the
"average" model achieved lower error on some of the sites in some years, but
each one had a higher mean absolute error and higher mean deviance (Figure 4).

![Site/year difference between the average model and all other models in absolute
error (**A.**) and deviance (**B.**). All testing years (2004-2013) are used here. 
None of the five other models provided a consistent improvement over the "average"
baseline, indicated by a horizontal line. Models' absolute error was generally
similar or larger than "average" model, with large outliers in both directions.
The other models' deviance was also generally higher than the "average"
baseline.](../figures/model_violins.png)

![Performance of all six models over time, according to three metrics. **A.**
Root mean square error (rmse) describes the distance between each model's point
estimates and the observed richness values. Note that the three environmental
models tend to show the largest errors. **B.** Deviance (-2 times the Gaussian
log-likelihood) describes the lack of fit of entire predictive distributions.
Note that the stack of single-species random forest SDMs shows much higher error
than the other models (since its predictions are too confident, see Fig. 1), and 
that the "naive" model's deviance grows relatively quickly (as its predictions
become broader and less informative, see Fig. 1). **C.** Coverage of a model's
95% confidence intervals indicates how often the observed values fall inside the
predicted range. For well-calibrated models, the coverage of the 95% confidence
intervals should be 95% (indicated by the black line). As in the previous panel,
the "naive" model's predictive distribution is too wide (capturing almost all of
the data) and the stacked SDM's predictive distribution is too narrow (missing
almost a third of the observed richness values by 2014).](../figures/performance_time.png)

Most models produced confidence intervals that were too narrow, indicating
overconfident predictions (Figure 5C). The random forest-based SDM stack was the
most overconfident model, with only `r numbers$rf_coverage_pct`% of observations
falling inside its 95% confidence intervals. This stacked SDM's narrow
predictive distribution also caused it to have notably higher deviance (Figure
5B) than the next-worst model, even though its point estimates were not
unusually bad in terms of RMSE (5A). As discussed elsewhere [@harris2015], this
overconfidence is a product of the assumption in stacked SDMs that errors in the
species-level predictions are independent. Because they avoided this assumption,
the GBM-based "macroecological" model and the mistnet JSDM had reasonably
well-calibrated uncertainty estimates (Figure 5B); as a result, their relative
performance was higher in terms of deviance than in terms of RMSE. The "naive"
model was the only model whose confidence intervals were too wide (Figure 5C),
which can be attributed to the rapid rate at which these intervals expand
(Figure 1).

Figure 6 shows each model's squared error, divided into consistent site-level
biases, versus errors that fluctuated year-to-year. Across all models, most of
the error came from errors in the site-level mean. The "average" model, which
was based on the site-level mean, had the lowest error in this regard. In
contrast, the three environmental models showed the largest biases at the site
level; this makes sense, given that they could not explicitly distinguish among
sites with similar climate, NDVI, and elevation.

![Each model's squared error can be divided into two parts. If a model
consistently over-estimates or under-estimates the richness at a given site,
this contributes to its error in predicting the site-level mean. The second
component is based on errors in predicting fluctuations in a site's richness
over time. Both components of the mean squared error were lower for the
single-site models than for the environmental models.](../figures/barcharts.png)

When the differences among observers were not explicitly accounted for, nearly
all metrics of fit worsened (Figure 7). The increased error resulted from a
small number of forecasts where observer turnover caused a large shift in the
reported richness values. The naive baseline was less sensitive to these shifts,
because it largely ignored the richness values reported by observers that had
retired by the end of the training period (Figure 1). The average model, which
gave equal weight to observations from the whole training period, showed a
larger decline in performance -- especially in terms of coverage. The
performance of the mistnet JSDM was notable here, because its prediction
intervals retained very good coverage in the presence of uncorrected
observer-based noise, which we attribute to the JSDM's ability to model this
variation with its latent variables.

![Accounting for differences between observers typically has a small impact on
accuracy (especially for the "naive" model), but occasonally changes a
prediction's accuracy by more than 10 species. These changes are beneficial, on
average.](../figures/observers.png)

>>>>

![Performance of six forecasting models for predicting species richness one year
(2004) and ten years into the future (2013). Plots show observed vs. predicted
values for species richness. Models were trained with data from 1982-2003.  In
general, the single-site models (**A**) outperformed the environmental models
(**B**). The accuracy of the predictions generally declined as the timescale of
the forecast was extended from 2004 to 2013.](../figures/scatter.png)

While all models generally performed well in absolute terms (Figure 3), none
consistently outperformed the "average" baseline (Figure 4). The auto-ARIMA was
generally the best-performing non-baseline model, but in many cases (67% of the
time), the auto.arima procedure selected a model with only an intercept term
(i.e. no autoregressive terms, no drift, and no moving average terms), making it
similar to the "average" model. All five alternatives to the "average" model
achieved lower error on some of the sites in some years, but each one had a
higher mean absolute error and higher mean deviance (Figure 4).

![Difference between the forecast error of models and the error of the average
baseline using both absolute error (**A.**) and deviance (**B.**). Differences
are taken for each site and testing year so that errors for the same forecast
are directly compared. The error of the average baseline is by definition zero
and is indicated by the horizontal gray line.  None of the five models provided
a consistent improvement over the average baseline. The absolute error of the
models was generally similar or larger than that of the "average" model, with
large outliers in both directions.  The deviance of the models was also
generally higher than the "average" baseline.](../figures/model_violins.png)

![Change in performance of the six forecasting models with the time scale of the
forecast (1-10 years into the future). **A.** Root mean square error (rmse; the
error in the point estimates) shows the three environmental models tending to
show the largest errors at all time scales and the models getting worse as they
forecast further into the future at approximately the same rate. **B.** Deviance
(lack of fit of the entire predictive distribution) shows the stacked species
distribution models with much higher error than other models and shows that the
"naive" model's deviance grows relatively quickly. **C.** Coverage of a model's
95% confidence intervals (how often the observed values fall inside the
predicted range; the black line indicates ideal performance) shows that the
"naive" model's predictive distribution is too wide (capturing almost all of the
data) and the stacked SDM's predictive distribution is too narrow (missing
almost a third of the observed richness values by
2014).](../figures/performance_time.png)

Most models produced confidence intervals that were too narrow, indicating
overconfident predictions (Figure 5C). The random forest-based SDM stack was the
most overconfident model, with only `r numbers$rf_coverage_pct`% of observations
falling inside its 95% confidence intervals. This stacked SDM's narrow
predictive distribution caused it to have notably higher deviance (Figure 5B)
than the next-worst model, even though its point estimates were not unusually
bad in terms of RMSE (5A). As discussed elsewhere [@harris2015], this
overconfidence is a product of the assumption in stacked SDMs that errors in the
species-level predictions are independent. The GBM-based "macroecological" model
and the mistnet JSDM had the best calibrated uncertainty estimates (Figure
5B)and therefore their relative performance was higher in terms of deviance than
in terms of RMSE. The "naive" model was the only model whose confidence
intervals were too wide (Figure 5C), which can be attributed to the rapid rate
at which these intervals expand (Figure 1).

Partitioning each model's squared error into consistent errors across sites and
errors that fluctuated year-to-year shows that the majority of the error comes
from errors in the site-level mean (Figure 6). The "average" model, which was
based on the site-level mean, had the lowest error in this regard. In contrast,
the three environmental models showed the largest biases at the site level; this
makes sense, given that they could not explicitly distinguish among sites with
similar climate, NDVI, and elevation.

![Partitioning of the squared error for each model into site and year
components. The site-level mean component shows consistent over or under
estimates of richness at a site across years. The annual fluctuation compoonent
shows errors in predicting fluctuations in a site's richness over time. Both
components of the mean squared error were lower for the single-site models than
for the environmental models.](../figures/barcharts.png)

Accounting for differences among observers generally improved measures of model
fit (Figure 7). Improvements primarily resulted from a small number of forecasts
where observer turnover caused a large shift in the reported richness
values. The naive baseline was less sensitive to these shifts, because it
largely ignored the richness values reported by observers that had retired by
the end of the training period (Figure 1). The average model, which gave equal
weight to observations from the whole training period, showed a larger decline
in performance when not accounting for observer effects -- especially in terms
of coverage. The performance of the mistnet JSDM was notable here, because its
prediction intervals retained good coverage even when not correcting for
observer differences, which we attribute to the JSDM's ability to model this
variation with its latent variables.

![Accounting for differences between observers typically has a small impact on
accuracy (especially for the "naive" model), but occasonally changes a
prediction's accuracy by more than 10 species. As a result the inclusion of
observer effects results in average improvements in overall model
fit.](../figures/observers.png)
